### 2019-06-08 2-Chapters/4-Chapter/IEEE_WCNC__2019__Paper__BMBM.git/IEEE_WCNC__2019__Paper__BMBM.tex
Small stuff

% In the context of Cognitive Radio \cite{Mitola99,Haykin05},
% Multi-Arm Bandit (MAB) algorithms \cite{Auer02,Auer02,Bubeck12} have been recently proposed as a potential solution for channel access in LPWA networks \cite{Bonnefoi18,Azari18,Bonnefoi17}.
We presented in Section~\ref{sec:4:firstModel} the impact of non-stationarity on the network performance using MAB algorithms is studied.
Low-cost algorithms following two well-known approaches, such as the Upper-Confidence Bound (\UCB{}) \cite{Auer02,Auer02}, and the Thompson Sampling (TS) algorithms \cite{Thompson33} have reported encouraging results.
When considering the application of MAB algorithms for slotted wireless protocols in a decentralized manner,
other recent directions of research first include theoretical analysis, like what we present in the next Chapter~\ref{chapter:5},
then realistic empirical simulations like in \cite{kumar2016two,kumar2017channel} or the previous Section~\ref{sec:4:gnuradio},
,
and finally applications to multi-hoping networks \cite{Mitton16,Toldov16} or other kinds of networks \cite{Wilhelmi19collaborative,Wilhelmi19potential}.
%
None of the mentioned works discuss in detail the impact of retransmissions on the performance of MAB learning algorithms as we do now.

@@

\paragraph{Outline.}

The rest of the section is organized as follows.
First the system model is introduced in Section~\ref{sub:43:model},
and our motivations are exposed in Section~\ref{sub:43:motivations}.
The proposed \UCB-based heuristics are presented in Section~\ref{sub:43:heuristics}, while the corresponding numerical results are shown in Section~\ref{sub:43:numExp}.
Finally, some conclusions are drawn later in Section~\ref{sub:43:conclusion}.

@@

It is well known \cite{Abramson1970,Roberts75}
that having a collision during an access time can be overcome by a retransmission procedure (this can take several retransmission attempts).
Our goal here is to obtain a mathematical approximation of the collision probability at the second transmission slot $p_{c1}$, as a function of the first collision probability $p_{c}$.

We make two approximations $\mathcal{H}_{1}$ and $\mathcal{H}_{2}$ defined as (they are hypotheses on which the rest of this section is built),
\begin{itemize}
	\item $\mathcal{H}_{1}$:
    The probability $p_{c1}$, is composed by the sum of two probabilities: i)
    the probability of colliding consecutively twice, \ie, the devices that collide at a given time slot and collide again when retransmitting their packets,
	and ii) the probability of collision among devices that did not collide in the same previous collision. Moreover, we suppose that the number of devices involved in a collision is small in comparison to the total number of devices.
	This is very realistic as a very small proportion of devices transmit at the same period, due to their low duty cycle.

@@

Moreover, from \eqref{eq:43:1} we define the probability $p_{cp}(n)$ that involves the collision of $n$ packets sent by each IoT device (for any $1\leq n \leq N-1$), during the first transmission slot, and is defined by the following equation
%
\begin{equation*}\label{eq:43:2}
	p_{cp}(n) = {N-1 \choose n} \; x^n \left(1-x\right)^{N-1-n}.
\end{equation*}

As explained above, if an IoT device experiences a collision at the first transmission, it proceeds for the retransmission of its packet after a random back-off interval.
We denote $p_{ca}$ the probability to have a collision with a packet involved in the previous collision.
Under the $\mathcal{H}_{1}$ assumption, the number of packets involved in the same previous collision remains very small in comparison to the total number of devices that may transmit during this time. In other words, this collision probability does not depend on previous retransmissions and is equal to $p_c$.
So, the probability that the same device's packet experiences again a collision at the second time slot is
%
\begin{equation}\label{eq:43:decomppc1}
	p_{c1} = p_{ca}+\left(1-p_{ca} \right)p_c.
\end{equation}

@@

Once again under $\mathcal{H}_{1}$, assuming that the number of devices involved in the first collision is small compared to $N-1$, the first $N_0 \ll N-1$ terms of the sum in \eqref{eq:43:sumpc2} are predominant. We derive,
%
\begin{equation}\label{eq:43:sumpca2}
	p_{ca} \simeq  1- \frac{1}{p_c}\sum_{n=1}^{N_0}{N-1 \choose n} x^n\left(1-x\right)^{N-1-n} \left( 1-\frac{1}{m}\right)^n.
\end{equation}

Moreover, for these terms, $n$ is small compared to $N-1$, and so $N-1-n$ can be approximated to $N-1$. Thus it gives,
%
\begin{equation}\label{eq:43:sumpca3}
	p_{ca} \simeq  1- \frac{\left(1-x\right)^{N-1}}{p_c}\sum_{n=1}^{N_0}{N-1 \choose n} x^n \left( 1-\frac{1}{m}\right)^n.
\end{equation}

@@

We interpret this fact in two different situations.
\begin{itemize}
	\item
	On the one hand, in a congested network, when devices suffer from a large probability of collision on their first transmission (\ie, $p_c$ is not so small), then $p_{c1}\simeq p_c$ and so devices cannot really hope to reduce their collision probabilities even if the use a different channel for retransmission.
	\item
	On the other hand, if $p_c$ is small enough, \ie, in a network not yet too congested, then our derivation shows that $p_{c1} > p_c$, meaning that the possible gain of retransmitting in a different channel that the one used for the first transmission can be large, in terms of collision probability (\eg, up-to $10\%$ in this experimental setting).
	In other words, when learning can be useful (small $p_c$), learning to retransmit in a different channel can have a large impact on the global collision rate,
	thus justifying our approach.
\end{itemize}

@@

\textbf{\textcolor{red}{Small warning about notations: local (device) time vs global (gateway) time}:}
%
In all the algorithms and notations presented in this Section, the time steps $t$ do not refer to the global time steps (as seen from the gateway), but rather the current number of uplink transmissions (or retransmissions) that was carried out by the device.
In other words, each device wakes up whenever it has to send some data, following its random emission process (here, we propose a Bernoulli process of small probability, \eg, $p=10^{-3}$), and when it woke up, it sends its packet, and will send again for at most $M$ times (\eg, $M=10$) until it receives an \Ack.

To be clear, once again, we repeat that the time steps $t$ in the following algorithms denote the number of time the device sent an uplink packet, and these uplink (re)transmissions can happen at different (real-world) times, more or less spread out in time, it does not matter to the learning part of this model and work.

>>>>

% In the context of Cognitive Radio \cite{Mitola99,Haykin05},
% Multi-Arm Bandit (MAB) algorithms \cite{Auer02,Auer02,Bubeck12} have been recently proposed as a potential solution for channel access in LPWA networks \cite{Bonnefoi18,Azari18,Bonnefoi17}.
We presented in Section~\ref{sec:4:firstModel} the impact of non-stationarity on the network performance using MAB algorithms is studied.
Low-cost algorithms following two well-known approaches, such as the Upper-Confidence Bound (\UCB{}) \cite{Auer02,Auer02}, and the Thompson Sampling (TS) algorithms \cite{Thompson33} have reported encouraging results.
When considering the application of MAB algorithms for slotted wireless protocols in a decentralized manner,
other recent directions of research first include theoretical analysis, like what we present in the next Chapter~\ref{chapter:5},
then realistic empirical simulations like in \cite{kumar2016two,kumar2017channel} or the previous Section~\ref{sec:4:gnuradio},
and finally applications to multi-hoping networks \cite{Mitton16,Toldov16} or other kinds of networks \cite{Wilhelmi19collaborative,Wilhelmi19potential}.
%
None of the mentioned works discuss in detail the impact of retransmissions on the performance of MAB learning algorithms as we do now.

@@

\textbf{Outline.}
%
The rest of the section is organized as follows.
First the system model is introduced in Section~\ref{sub:43:model},
and our motivations are exposed in Section~\ref{sub:43:motivations}.
The proposed \UCB-based heuristics are presented in Section~\ref{sub:43:heuristics}, while the corresponding numerical results are shown in Section~\ref{sub:43:numExp}.
Finally, some conclusions are drawn later in Section~\ref{sub:43:conclusion}.

@@

It is well known \cite{Abramson1970,Roberts75}
that having a collision during an access time can be overcome by a retransmission procedure (this can take several retransmission attempts).
Our goal here is to obtain a mathematical approximation of the collision probability at the second transmission slot $p_{c1}$, as a function of the first collision probability $p_{c}$.
%
We make two approximations $\mathcal{H}_{1}$ and $\mathcal{H}_{2}$ defined as (they are hypotheses on which the rest of this section is built),
\begin{itemize}
	\item $\mathcal{H}_{1}$:
    The probability $p_{c1}$, is composed by the sum of two probabilities: i)
    the probability of colliding consecutively twice, \ie, the devices that collide at a given time slot and collide again when retransmitting their packets,
	and ii) the probability of collision among devices that did not collide in the same previous collision. Moreover, we suppose that the number of devices involved in a collision is small in comparison to the total number of devices.
	This is very realistic as a very small proportion of devices transmit at the same period, due to their low duty cycle.

@@

Moreover, from \eqref{eq:43:1} we define the probability $p_{cp}(n)$ that involves the collision of $n$ packets sent by each IoT device (for any $1\leq n \leq N-1$), during the first transmission slot, and is defined by $p_{cp}(n) = {N-1 \choose n} \; x^n \left(1-x\right)^{N-1-n}$.
%
As explained above, if an IoT device experiences a collision at the first transmission, it proceeds for the retransmission of its packet after a random back-off interval.
We denote $p_{ca}$ the probability to have a collision with a packet involved in the previous collision.
Under the $\mathcal{H}_{1}$ assumption, the number of packets involved in the same previous collision remains very small in comparison to the total number of devices that may transmit during this time. In other words, this collision probability does not depend on previous retransmissions and is equal to $p_c$.
So, the probability that the same device's packet experiences again a collision at the second time slot is
%
\begin{equation}\label{eq:43:decomppc1}
	p_{c1} = p_{ca}+\left(1-p_{ca} \right)p_c.
\end{equation}

@@

Once again under $\mathcal{H}_{1}$, assuming that the number of devices involved in the first collision is small compared to $N-1$, the first $N_0 \ll N-1$ terms of the sum in \eqref{eq:43:sumpc2} are predominant. We derive $p_{ca} \simeq  1- \frac{1}{p_c}\sum_{n=1}^{N_0}{N-1 \choose n} x^n\left(1-x\right)^{N-1-n} \left( 1-\frac{1}{m}\right)^n$.
%
Moreover, for these terms, $n$ is small compared to $N-1$, and so $N-1-n$ can be approximated to $N-1$. Thus it gives,
%
\begin{equation}\label{eq:43:sumpca3}
	p_{ca} \simeq  1- \frac{\left(1-x\right)^{N-1}}{p_c}\sum_{n=1}^{N_0}{N-1 \choose n} x^n \left( 1-\frac{1}{m}\right)^n.
\end{equation}

@@

We interpret this fact in two different situations:
\begin{itemize}
	\item
	On the one hand, in a congested network, when devices suffer from a large probability of collision on their first transmission (\ie, $p_c$ is not so small), then $p_{c1}\simeq p_c$ and so devices cannot really hope to reduce their collision probabilities even if the use a different channel for retransmission.
	\item
	On the other hand, if $p_c$ is small enough, \ie, in a network not yet too congested, then our derivation shows that $p_{c1} > p_c$, meaning that the possible gain of retransmitting in a different channel that the one used for the first transmission can be large, in terms of collision probability (\eg, up-to $10\%$ in this experimental setting).
	In other words, when learning can be useful (small $p_c$), learning to retransmit in a different channel can have a large impact on the global collision rate,
	thus justifying our approach.
\end{itemize}

@@

\begin{leftbar}[warningbar]  % XXX leftbar warningbar, comment if needed
	\textbf{\textcolor{red}{Small warning about notations: local (device) time vs global (gateway) time}:}
	%
	In all the algorithms and notations presented in this Section, the time steps $t$ do not refer to the global time steps (as seen from the gateway), but rather the current number of uplink transmissions (or retransmissions) that was carried out by the device.
	Each device wakes up whenever it has to send some data, following its random emission process (here, we propose a Bernoulli process of small probability, \eg, $p=10^{-3}$), and when it woke up, it sends its packet, and will send again for at most $M$ times (\eg, $M=10$) until it receives an \Ack.
	%
	In other words, the time steps $t$ in the following algorithms denote the number of time the device sent an uplink packet, and these uplink (re)transmissions can happen at different (real-world) times, more or less spread out in time, it does not matter to the learning part of this model and work.
\end{leftbar}  % XXX leftbar warningbar, comment if needed

### 2019-06-08 2-Chapters/5-Chapter/ALT_2018__MPBandits.git/mpBandits.tex
Small stuff

% FIXME
% \begin{framed}
  \textbf{\textcolor{red}{Warning}.}
  The regret lower bound given in \cite{Besson2018ALT} is not applicable to any algorithm,
  as it was discovered by Etienne Boursier and Vianney Perchet, as they explain in Section~2.4 in \cite{BoursierPerchet18}.
  The proof we gave in the Appendix of our paper \cite{Besson2018ALT} was wrong in just one step,
  but the results stated in this section are now correct,
  thanks to the modifications proposed by Emilie Kaufmann and Abbas Mehrabian in Appendix~E of \cite{KaufmannAbbas19}.
  The cost for this modification is to restrict the lower-bound to a smaller class of algorithms, which \emph{should} contain \RhoRand, \RandTopM{} and \MCTopM, but not the two algorithms proposed SIC-MMAB from \cite{BoursierPerchet18} and Multiplayer Explore-then-Commit (M-ETC) from \cite{KaufmannAbbas19}.
  Proving that the proposed algorithms are in this class is left as an open question.
% \end{framed}
% FIXME

@@

We introduce additional notations in the following definition.

\begin{definition}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
  % [$T_k(T)$ and $\cC_k(T)$]
  \label{def:5:nbSelections_nbCollisions}
  % \label{def:5:nbSelections}
  Let $T^j_k(T) := \sum_{t=1}^T \indic(A^j(t) = k)$,
  and denote $T_k(T) := \sum_{j=1}^M T^j_k(T)$ the \emph{number of selections} of arm $k\in[K]$ by any player $j\in[M]$, up to time $T$.
  % and $\cC_k(T)$ be the number of collisions on arm $k$ up-to time $T$.
% \end{definition}

>>>>

\begin{leftbar}[warningbar]  % XXX leftbar warningbar, comment if needed
  % FIXME
  \textbf{\textcolor{red}{Warning}.}
  %
  The regret lower bound given in \cite{Besson2018ALT} is not applicable to any algorithm,
  as it was discovered by Etienne Boursier and Vianney Perchet, as they explain in Section~2.4 in \cite{BoursierPerchet18}.
  The proof we gave in the Appendix of our paper \cite{Besson2018ALT} was wrong in just one step,
  but the results stated in this section are now correct,
  thanks to the modifications proposed by Emilie Kaufmann and Abbas Mehrabian in Appendix~E of \cite{KaufmannAbbas19}.
  The cost for this modification is to restrict the lower-bound to a smaller class of algorithms, which \emph{should} contain \RhoRand, \RandTopM{} and \MCTopM, but not the two algorithms proposed SIC-MMAB from \cite{BoursierPerchet18} and Multiplayer Explore-then-Commit (M-ETC) from \cite{KaufmannAbbas19}.
  Proving that the proposed algorithms are in this class is left as an open question.
\end{leftbar}  % XXX leftbar warningbar, comment if needed

@@

We introduce additional notations in the following definition.
%
\begin{definition}\label{def:5:nbSelections_nbCollisions}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
  % [$T_k(T)$ and $\cC_k(T)$]
  % \label{def:5:nbSelections}
  Let $T^j_k(T) := \sum_{t=1}^T \indic(A^j(t) = k)$,
  and denote $T_k(T) := \sum_{j=1}^M T^j_k(T)$ the \emph{number of selections} of arm $k\in[K]$ by any player $j\in[M]$, up to time $T$.
  % and $\cC_k(T)$ be the number of collisions on arm $k$ up-to time $T$.
% \end{definition}

### 2019-06-08 2-Chapters/5-Chapter/chapter5.tex
Small stuff

% ----------------------------------------------------------------------------
\subsection{Inspirations from real-world demonstrations to validate some models}
\label{sub:5:USRPdemos}

@@

In \cite{darak2016bayesian,Darak16}, the authors study the same model and focussed on the \RhoRand{} policy, which was the state-of-the-art back in $2016$, combined with \UCB{} or \klUCB, as well as an extension using Bayes-UCB from \cite{Kaufmann12BUCB}.
Their idea is to use the same skeleton as \RhoRand, that is to assign a \emph{rank} to each player, and make players change their rank after any collision. But instead of selecting a new rank uniformly at random among $[M]$, they proposed to use a ``second-stage'' learning policy\footnote{This is a very natural idea: use a bandit algorithm to balance the exploration/exploitation aspect of rank selection, instead of a random hoping.}, based on Bayes-UCB, to (try to) learn the best rank while still exploring each rank from time to time.
This algorithm is implemented in SMPyBandits and named \rhoLearn,
% \footnote{See \href{https://smpybandits.github.io/docs/PoliciesMultiPlayers.rhoLearn.html}{\texttt{SMPyBandits.github.io/docs/PoliciesMultiPlayers.rhoLearn.html}} for its documentation.},
where it can use Bayes-UCB or any of the $65$ or more bandit algorithms available in the library.
Even if no theoretical guarantee backs up this idea of second-stage learning with Bayes-UCB, empirical simulations found that it can be efficient.
We illustrate this in Table~\ref{table:5:comparisonRhoRandRhoLearn} below, where we consider the same problem as described above in Section~\ref{sub:5:withoutSensing} (Table~\ref{table:5:meanRegretSimulationsNoSensing}).
The first \rhoLearn uses \klUCB{}, Bayes-UCB or Exp3 for the second-stage learning,
and we include both the centralized multi-play version of \klUCB{} and \MCTopM-\klUCB{} for comparison.

>>>>

% ----------------------------------------------------------------------------
\subsection{Inspirations from Cognitive Radio Proof-of-Concepts}
\label{sub:5:USRPdemos}

@@

In \cite{darak2016bayesian,Darak16}, the authors study the same model and focussed on the \RhoRand{} policy, which was the state-of-the-art back in $2016$, combined with \UCB{} or \klUCB, as well as an extension using Bayes-UCB from \cite{Kaufmann12BUCB}.
Their idea is to use the same skeleton as \RhoRand, that is to assign a \emph{rank} to each player, and make players change their rank after any collision. But instead of selecting a new rank uniformly at random among $[M]$, they proposed to use a ``second-stage'' learning policy\footnote{This is a very natural idea: use a bandit algorithm to balance the exploration/exploitation aspect of rank selection, instead of a random hoping.}, based on Bayes-UCB, to (try to) learn the best rank while still exploring each rank from time to time.
This algorithm is implemented in SMPyBandits and named \rhoLearn,
% \footnote{See \href{https://smpybandits.github.io/docs/PoliciesMultiPlayers.rhoLearn.html}{\texttt{SMPyBandits.github.io/docs/PoliciesMultiPlayers.rhoLearn.html}} for its documentation.},
where it can use Bayes-UCB or any of the $65$ or more bandit algorithms available in the library.
Even if no theoretical guarantee backs up this idea of second-stage learning with Bayes-UCB, empirical simulations found that it can be efficient.
We illustrate this in Table~\ref{table:5:comparisonRhoRandRhoLearn} below, where we consider the same problem as described above in Section~\ref{sub:5:withoutSensing} (Table~\ref{table:5:meanRegretSimulationsNoSensing}).
The first \rhoLearn uses \klUCB{}, Bayes-UCB or Exp3 for the second-stage learning,
and we include both the centralized multiple-play version of \klUCB{} and \MCTopM-\klUCB{} for comparison.
